FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-01-24 epss_marmo.tex

\subsection{FITS format}
FITS\citep{fitsorig,fitsver3,fitsoffice} stands for 'Flexible Image Transport System'.
It is an open digital standard, defined for data acquisition and archiving in astronomical
observatories back in the late 70's, and it is used for spatial telescope data.
In fact, FITS is the IAU-approved standard format for astronomical
data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard implemented formats in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html} (VO).
Finally, FITS format is compatible with the Planetary Data System (PDS)
3\footnote{https://pds.nasa.gov/pds4/about/} and
4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml} archiving
specifications and supported by a large number of open libraries and software tools.

FITS is already able to propose standard formatting for some data products quite
common by now in planetary surface investigations.
In particular, Multi-Extension FITS (MEF) schema proposes an easy way
to store inhomogeneous digital information (reflectance, calibration data, vector
table data, etc.) in the same file each with relative metadata, as well as
multi-detector imagery (e.g. from HiRISE \citep{hirise}) or hyperspectral cubes
(e.g. from CRISM \citep{crism} or OMEGA \citep{omega}).
FITS has been already chosen to distribute, e.g.,
Hayabusa\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
Akatsuki\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
and some of the Dawn\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
To be efficiently used in planetary surface investigations, FITS metadata must
be extended in order to take into account the size of the reference body.
Planetary body shapes and orientations are well defined and standardized by the
International Astronomical Union (IAU) Working Group on Cartographic
Coordinates and Rotational Elements (WGCCRE).
This group reports triennially on the preferred rotation rate, spin axis, prime meridian,
and reference surface for planets and satellites which helps ensure that
cartographic endeavors are effectively comparable \citep{wgccre}.
In the framework of the VESPA [12, 13] component of the
Europlanet 2020 project an extension to FITS meta-
data (GeoFITS) has been proposed [14].

\subsubsection{Planetary Surface proposed convention}

>>>>

\subsection{FITS format}
FITS\footnote{https://fits.gsfc.nasa.gov/} \citep{fitsorig,fitsver3} stands for
'Flexible Image Transport System'.
It is an open digital standard, defined for data acquisition and archiving in astronomical
observatories back in the late 70's, and it is used for spatial telescope data.
In fact, FITS is the IAU-approved standard format for astronomical
data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard implemented formats in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html} (VO).
Finally, FITS format is compatible with the Planetary Data System (PDS)
3\footnote{https://pds.nasa.gov/pds4/about/} and
4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml} archiving
specifications and supported by a large number of open libraries and software tools.

FITS is already able to propose standard formatting for some data products quite
common by now in planetary surface investigations.
In particular, Multi-Extension FITS (MEF) schema proposes an easy way
to store inhomogeneous digital information (reflectance, calibration data, vector
table data, etc.) in the same file each with relative metadata, as well as
multi-detector imagery (e.g. from HiRISE \citep{hirise}) or hyperspectral cubes
(e.g. from CRISM \citep{crism} or OMEGA \citep{omega}).
FITS has been already chosen to distribute, e.g.,
Hayabusa\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
Akatsuki\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
and some of the Dawn\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
To be efficiently used in planetary surface investigations, FITS metadata must
be extended in order to take into account the size of the reference body.
Planetary body shapes and orientations are well defined and standardized by the
International Astronomical Union (IAU) Working Group on Cartographic
Coordinates and Rotational Elements (WGCCRE).
This group reports triennially on the preferred rotation rate, spin axis, prime meridian,
and reference surface for planets and satellites which helps ensure that
cartographic endeavors are effectively comparable \citep{wgccre}.
In the framework of the VESPA\footnote{http://europlanet-vespa.eu/} \citep{vespapss}
component of the EuroplanetRI2020 project an extension to FITS meta-
data (GeoFITS) has been proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention}.

\subsubsection{Planetary Surface proposed convention}
When willing to use FITS format for Planetary Surfaces the issue of rigorous
geographical representation must be addressed.
Planetary surface researchers for their mapping and analysis needs, commonly use
Geographical Information Systems (GIS).
In general, GIS applications excel in data interoperability even
though some have historically been anchored to Earth's spatial description.
FITS World Coordinate System representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html.}
simplifies the spatial coordinate description with respect to historical terrestrial
references.
There is no need for oblique projection definitions \citep{Snyder1987,calagreisenI}
as projection parameters (projection center or reference point)
describe the difference between the Mercator and
Transverse Mercator or Simple Cylindrical and
Equirectangular map projections.
There is no need for East or West positive longitude definitions as the pixel
conversion in world coordinates is set by a common
oriented matrix. 
However, if still needed, FITS does offer the possibility to fully describe those
same parameters using alternative WCS definitions (meters, degrees,
East and West longitude, etc.)\citep{greisencalaII}.
Nevertheless, WCS does not record information about the body shape and
orientation, and reference surface.
